Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-13 paper/paper.tex
Update numbers

\begin{figure}[h]
  \begin{tabular}{ |l |  c  | c | c| }
    \hline
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    Syntactic typing         &  83 & =   & = \\
    Untyped reduction        & 344 & =   & = \\
    Neutral and normal forms & -   & 273 & - \\
    Logical relation         & 338 & 430 & - \\
    Semantic soundness       & 192 & 211 & - \\
    Syntactic soundness      & -   & -   & 629 \\
    \hline
    Total                    & 957 & 1341 & 1056 \\
    \hline
  \end{tabular}
  \caption{Nonblank, noncomment lines of code of the Coq Development. The marker = indicates that the line count is the same as the column to the
left. The marker - indicates the file does not
contribute to the total.}
  \label{fig:linecount}
\end{figure}

To demonstrate the scale of our proof scripts, Figure~\ref{fig:linecount}
shows the number of non-blank, non-comment lines of code\footnote{calculated
  by the \texttt{tokei} tool, available from
  \url{https://github.com/XAMPPRocky/tokei}.} for each file of our development,
including the base consistency proof from Section~\ref{sec:logreldep} and
\ref{sec:logrelproof} and the extension to $\beta$-normalization from
Section~\ref{sec:extension}.  For comparison, we have also proven syntactic
type soundness through preservation
\footnote{\dotv{syntactic\_soundness.v}{subject\_reduction}} and
progress\footnote{\dotv{syntactic\_soundness.v}{wt\_progress}}.

The $\beta\eta$-normalization proof from Section~\ref{sec:betaeta} comprises
1568 lines of non-blank, non-comment lines of code. We choose not to include
it in the chart, because of slight differences in lemma dependencies for
untyped reduction and normal forms that make the comparison less
informative. However, when compared to the $\beta$-normalization extension,
the $\beta\eta$ extension has the same line count in the definition of the
logical relation and the semantic soundness proof.

The Autosubst 2 tool takes our 13 line syntax specification, written in
higher-order abstract syntax, and generates the Coq syntax specification,
renaming and substitution functions, and lemmas and tactics that allow
reasoning about those functions. The auto-generated syntax file (291 LOC)
and other Autosubst library files are also not included in the figure.

\subsection{Axioms}
Our Coq development assumes two axioms: functional extensionality
and propositional extensionality. The former is also required by
the Autosubst 2 libraries. Both axioms are known to be consistent
with Coq's metatheory.
%
These axioms bridge the gap between our mechanization and our informal
proofs. For example, in set theory, to show that two sets $[[S0]]$ and
$[[S1]]$ are equal, it suffices to show the extensional property that
$\forall x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
occasionally in our presented proofs.
%
However, in Coq, sets of terms ($[[PowerSet STm]]$) are encoded as the 
type \texttt{tm
  -> Prop}, a predicate over \lang{} terms.  In axiom-free Coq, predicates do
not come with the extensionality property. Given two predicates $P$ and $Q$,
we cannot conclude that $P = Q$ when given a proof of
$\forall x, P(x) \iff Q(x)$. But this is exactly the statement of 
predicate extensionality, an immediate corollary from functional
extensionality and propositional extensionality.

@@

As discussed, the logical relations $[[Interp I i A S]]$ in
Figures~\ref{fig:logrel} and \ref{fig:logrelopen} are encoded as inductive
definitions indexed by sets of terms representing the semantic inhabitants of
$[[A]]$. In the Coq mechanization, we represent sets of terms as predicates
over terms asserting whether the term is in the set, so $[[S]]$ has type
\texttt{tm -> Prop}, and $[[I]]$ has type \texttt{forall j, j < i -> tm -> Prop},
representing the set of types in universes strictly smaller than $[[i]]$.
Because the overall inductive definition represents a functional relation
between a type at a universe level and its interpretation as a set of terms,
it too lives in \texttt{Prop}.

@@

Further details of the IR encoding can be found in our Agda mechanization of
the consistency of \lang. Due to the lack of automation support, proofs about
syntactic properties of terms and proofs using well-founded induction on levels
are written manually. Even so, the total number of nonblack, noncomment lines
of code remains small at below 1200 lines, and functional extensionality is
still the only axiom used.

@@

There are other mechanized developments for dependently typed systems that
only involve properties that are derivable through syntactic means. For
example, \citet{coqcoqcorrect2019} prove the correctness of a type checker
for the Predicative, Cumulative Calculus of Inductive Constructions (PCUIC),
Coq's core calculus, assuming the strong normalization property of the object
language. \citet{weirich:systemd} define System D, a core calculus of
dependent Haskell, and prove the syntactic type soundness of the type
system. Because System D includes nontermination, they proved
the consistency of definitional equality from the confluence of 
parallel reduction.

>>>>

\begin{figure}[h]
  \centering
  \begin{tabular}{ | l |  c  | c | c | c | }
    \hline
    & Consistency & Normalization & Safety\\
    \hline
    Libraries            &  15 & $=\joinrel=$ & $=\joinrel=$ \\
    Syntactic typing     &  87 & $=\joinrel=$ & $=\joinrel=$ \\
    Untyped reduction    & 345 & $=\joinrel=$ & $=\joinrel=$ \\
    Neutral/normal forms & --- & 287  & ---  \\
    Logical relation     & 295 & 385  & ---  \\
    Semantic soundness   & 195 & 215  & ---  \\
    Syntactic soundness  & --- & ---  & 655  \\
    \hline
    Total                & 934 & 1331 & 1099 \\
    \hline
  \end{tabular}
  \caption{Nonblank, noncomment lines of code of the Coq development.
    The $=\joinrel=$ marker indicates that the line count is the same as the column to the left.
    The --- marker indicates the file does not contribute to the total.}
  \label{fig:linecount}
\end{figure}

To demonstrate the scale of our proof scripts, Figure~\ref{fig:linecount}
shows the number of non-blank, non-comment lines of code%
  \footnote{calculated by \texttt{tokei}:
    \url{https://github.com/XAMPPRocky/tokei} \hfill}
for each file of our development,
including the base consistency proof from Sections~\ref{sec:logreldep} and
\ref{sec:logrelproof}, along with the extension to $\beta$-normalization from
Section~\ref{sec:extension}. For comparison, we have also proven syntactic
type safety through preservation
\footnote{\dotv{syntactic\_soundness.v}{subject\_reduction}} and
progress\footnote{\dotv{syntactic\_soundness.v}{wt\_progress}}.

The $\beta\eta$-normalization proof from Section~\ref{sec:betaeta} comprises
1549 lines of non-blank, non-comment lines of code. We choose not to include
it in the chart, because of slight differences in lemma dependencies for
untyped reduction and normal forms that make the comparison less
informative. However, when compared to the $\beta$-normalization extension,
the $\beta\eta$ extension has the same line count in the definition of the
logical relation and the semantic soundness proof.

The Autosubst 2 tool takes our 16-line syntax specification, written in
higher-order abstract syntax, and generates the Coq syntax specification,
renaming and substitution functions, and lemmas and tactics that allow
reasoning about those functions. The autogenerated syntax file and other
Autosubst library files (516 LOC total) are not included in the figure.

\subsection{Axioms}

Our Coq development assumes two axioms: functional extensionality
and propositional extensionality. The former is also required by
the Autosubst 2 libraries. Both axioms are known to be consistent
with Coq's metatheory.
%
These axioms bridge the gap between our mechanization and our informal
proofs. For example, in set theory, to show that two sets $[[S0]]$ and
$[[S1]]$ are equal, it suffices to show the extensional property that
$\forall x, x \in [[S0]] \iff x \in [[S1]]$. We use this property in a few
proofs, notably for inversion (Lemma~\ref{lemma:interpinv}) and functionality
(Lemma~\ref{lemma:logreldeter}), where we deal with equalities of
the interpretations, which are sets of terms.

In the Coq mechanization, sets of terms $[[PowerSet STm]]$ are encoded as the
predicates over terms \texttt{tm -> Prop} asserting whether the term is in the
set. We want to be able to show that two predicates $P$ and $Q$ are equal when
$\forall x, P(x) \iff Q(x)$ holds, but this predicate extensionality property
does not hold in axiom-free Coq, and requires both functional and propositional
extensionality.

@@

As discussed, the logical relations $[[Interp I i A S]]$ in
Figures~\ref{fig:logrel} and \ref{fig:logrelopen} are encoded as inductive
definitions indexed by sets of terms representing the semantic inhabitants of
$[[A]]$. $[[S]]$ therefore has type \texttt{tm -> Prop}, while $[[I]]$ has type
\texttt{forall j, j < i -> tm -> Prop}, representing a set of types in
universes strictly smaller than $[[i]]$. Because the overall inductive
definition represents a functional relation between a type at a universe level
and its interpretation as a set of terms, it too lives in \texttt{Prop}.

@@

Further details of the IR encoding can be found in our Agda mechanization of
the consistency of \lang. Due to the lack of automation support, proofs about
syntactic properties of terms and proofs using well-founded induction on levels
are written manually. Even so, the total number of nonblack, noncomment lines
of code remains small at below 1300 lines, and functional extensionality is
still the only axiom used.

@@

There are other mechanized developments for dependently typed systems that
only involve properties that are derivable through syntactic means. For
example, \citet{coqcoqcorrect2019} prove the correctness of a type checker
for the Predicative, Cumulative Calculus of Inductive Constructions (PCUIC),
Coq's core calculus, assuming the strong normalization property of the object
language. \citet{weirich:systemd} define System D, a core calculus of
dependent Haskell, and prove the syntactic type safety of the type
system. Because System D includes nontermination, they proved
the consistency of definitional equality from the confluence of 
parallel reduction.
